Figure 2 - Overview heatmaps for IP-MS and ChIP

### 2024-07-09 ms_figure_2_overview-heatmaps.Rmd
Lift out IP-MS heatmap functions into scripts, move HOT IP-MS heatmap to Supp Fig 2

---
title: "Figure 2 - Overview heatmaps for IP-MS and ChIP"
author: "Charlotte Soneson, Michael Stadler, Merle Skribbe"
date: "`r Sys.Date()`"
output: 
    html_document:
        theme: united
        toc: true
        toc_float: true
        code_folding: hide
params:
    rds150: "data/ipms_150_sce.rds"
    rds500: "data/ipms_500_sce.rds"
    complexes: "data/complexes.json"
    idmap: "data/id_mapping_table.txt"
    baitclass: "data/ipms_bait_class.txt"
    dbdtxt: "data/TF_DBD_annotation.txt"
    peakcsv: "data/fused_peaks_filtered.csv.gz"
    peakenr: "data/fused_peaks_filtered_enrichments.csv.gz"
    peakatacfile: "data/fused_peaks_filtered_counts-external_atac.csv.gz"
    peakchipfile: "data/fused_peaks_filtered_counts-external_chip.csv.gz"
    gtf: "reference/Schizosaccharomyces_pombe.ASM294v2.55.gtf"
---

@@

## Source scripts with required helper functions and settings
source("params/plot_settings.R")
source("params/get_testres_function.R")
source("params/ipms_params.R")
source("params/mapping_functions.R")

@@

```{r heatmap-helper}
filterForHeatmap <- function(tstatmat, adjpmat, logfcmat, adjpThreshold, 
                             logfcThreshold, maxNbrTargets) {
    ## Filter - require significant in at least one comparison
    tstatfilt <- tstatmat
    tstatfilt[adjpmat >= adjpThreshold] <- NA
    tstatfilt[logfcmat <= logfcThreshold] <- NA
    # tstatfilt[tstatfilt <= 0] <- NA

    ## Some experiments pull down a large number of proteins
    # print(table(colSums(!is.na(tstatfilt))))
    ## -> remove proteins that are only being pulled down in these experiments
    keep <- which(colSums(!is.na(tstatfilt)) <= maxNbrTargets)
    tstatfilt <- tstatfilt[rowSums(!is.na(tstatfilt[, keep])) > 0, ]
    tstatfilt[is.na(tstatfilt)] <- 0
    tstatfilt <- tstatfilt[rowSums(tstatfilt) > 0, ]

    ## Subset the t-statistic matrix to only the retained proteins
    stopifnot(colnames(tstatfilt) == colnames(tstatmat))
    tstatmat <- tstatmat[rownames(tstatfilt), colnames(tstatfilt)]
    colnames(tstatmat) <- .getSimplifiedComparison(colnames(tstatmat))

    list(tstats = tstatmat, tstatsfilt = tstatfilt)
}

makeComplexAnnotation <- function(tstatmat, complexes, idmap,
                                  show_legend = TRUE) {
    rowannot <- data.frame(Complex = rep(NA, nrow(tstatmat)),
                           row.names = rownames(tstatmat))
    rowcols <- list(Complex = c())
    for (cmplx in names(complexes)) {
        rowannot$Complex[.getPomBaseIdFromProtein(rownames(rowannot), 
                                                  idmap = idmap) %in% 
                             complexes[[cmplx]]$pombase_ids] <- cmplx
        rowcols$Complex[cmplx] <- complexes[[cmplx]]$color
    }
    rowAnnotation(df = rowannot, 
                  col = rowcols, 
                  annotation_name_gp = gpar(fontsize = fl),
                  simple_anno_size_adjust = TRUE,
                  na_col = bg_color,
                  show_annotation_name = FALSE,
                  show_legend = show_legend,
                  annotation_legend_param = list(title_gp = gpar(fontsize = fl)))
}

makeColLabels <- function(tstatmat, colLabel, fontsize) {
    columnAnnotation(
        TFnames = anno_mark(which = "column", side = "bottom",
                            at = match(colLabel, colnames(tstatmat)), 
                            labels = .getProteinNameFromSimplifiedComparison(colLabel),
                            labels_gp = gpar(fontsize = fontsize)))
}

makeHeatmapCol <- function(tstatmat, stringency = "high") {
    if (stringency == "high") {
        circlize::colorRamp2(
            breaks = seq(3.0, 13.0, length.out = 64),
            colors = colorRampPalette(binary_heatmap_colors[c("FALSE", "TRUE")])(64))
    } else {
        circlize::colorRamp2(
            breaks = seq(1.5, 13.0, length.out = 64),
            colors = colorRampPalette(binary_heatmap_colors[c("FALSE", "TRUE")])(64))
    }
}

reorderColsByRows <- function(tstat) {
    ## Row names should be protein names, column names simplified comparison names
    ord <- match(rownames(tstat), .getProteinNameFromSimplifiedComparison(colnames(tstat)))
    ord <- ord[!is.na(ord)]
    ord <- c(ord, setdiff(seq_len(ncol(tstat)), ord))
    tstat[, ord]
}

makeHeatmapData <- function(sce, adjpthr, log2fcthr, conc) {
    tc <- .getTestCols(sce, adjp_cutoff = adjpthr, logfc_cutoff = log2fcthr)

    tstats <- filterForHeatmap(tstatmat = tc$tstat, adjpmat = tc$adjp, 
                               logfcmat = tc$logfc, logfcThreshold = log2fcthr,
                               adjpThreshold = adjpthr, maxNbrTargets = Inf)
    tstat <- tstats$tstats

    ## tstats$tstatsfile: protein x experiment
    ## Number of TFs that are pulled down by each bait
    all_tfs <- setdiff(.getProteinNameFromComparison(colnames(tstats$tstatsfilt)), 
                       c("Untagged", "untagged", NA))
    nbrInt <- colSums(tstats$tstatsfilt[rownames(tstats$tstatsfilt) %in% all_tfs, ] > 0, 
                      na.rm = TRUE)
    names(nbrInt) <- .getSimplifiedComparison(names(nbrInt))

    ## Split by family - return a matrix with several columns: 
    ## #pulled down TFs (including itself), #pulled down TFs (excluding itself),
    ## #pulled down TFs from the same family (excluding itself), 
    ## #pulled down TFs from another family
    tmptf <- tstats$tstatsfilt[match(.getProteinNameFromComparison(colnames(tstats$tstatsfilt)), 
                                     rownames(tstats$tstatsfilt)), ]
    rownames(tmptf) <- .getProteinNameFromComparison(colnames(tstats$tstatsfilt))
    dbdtf <- data.frame(PomBaseID = .getPomBaseIdFromComparison(colnames(tstats$tstatsfilt), 
                                                                idmap = idmap),
                        bait = .getProteinNameFromComparison(colnames(tstats$tstatsfilt)),
                        DBD_class = NA)
    for (i in seq_len(nrow(dbdtf))) {
        lookfor <- dbdtf$PomBaseID[i]
        if (!(lookfor %in% c("untagged", "Untagged"))) {
            dbdtf$DBD_class[i] <- dbd$DBD_class[match(lookfor, dbd$PomBaseID)]
        } else {
            dbdtf$DBD_class[i] <- "no_class"
        }
    }
    stopifnot(dbdtf$bait == .getProteinNameFromComparison(colnames(tmptf)))
    nbrIntMat <- do.call(dplyr::bind_rows, lapply(seq_along(colnames(tmptf)), function(i) {
        expr <- colnames(tmptf)[i]
        fam <- dbdtf$DBD_class[i]
        data.frame(experiment = expr, 
                   bait = dbdtf$bait[i], 
                   dbdfam = fam, 
                   TFs = paste(rownames(tmptf)[-i][which(tmptf[-i, expr] > 0)],
                               collapse = ","),
                   nbrTFs = sum(tmptf[, expr] > 0, na.rm = TRUE),
                   nbrTFs_wobait = sum(tmptf[-i, expr] > 0, na.rm = TRUE),
                   nbrTFs_samefam = sum(tmptf[-i, expr] > 0 & 
                                            dbdtf$DBD_class[-i] == fam, 
                                        na.rm = TRUE),
                   nbrTFs_difffam = sum(tmptf[-i, expr] > 0 & 
                                            dbdtf$DBD_class[-i] != fam, 
                                        na.rm = TRUE)
        )
    }))

    ## Split experiments into three groups
    if (conc == 150) {
        stopifnot(all(colnames(tstat) == .getSimplifiedComparison(colnames(tstats$tstatsfilt))))
        tstat_no_bait <- 
            tstat[, colSums(tstats$tstatsfilt) == 0 | 
                      (.getOrigBaitNameFromComparison(colnames(tstats$tstatsfilt), idmap = idmap) %in% 
                      baitclass$Gene_name[baitclass$class == "nobait"])]
        tstat_150only <- 
            tstat[, colSums(tstats$tstatsfilt) > 0 &
                      (.getOrigBaitNameFromComparison(colnames(tstats$tstatsfilt), idmap = idmap) %in% 
                      baitclass$Gene_name[baitclass$class == "150only"])]
        tstat_150and500 <- 
            tstat[, colSums(tstats$tstatsfilt) > 0 &
                      (.getOrigBaitNameFromComparison(colnames(tstats$tstatsfilt), idmap = idmap) %in% 
                      baitclass$Gene_name[baitclass$class == "150and500"])]

        proteins_150only <- .getProteinNameFromSimplifiedComparison(colnames(tstat_150only))
        stopifnot(all(proteins_150only %in% rownames(tstat)))
        tstat_150and500 <- tstat_150and500[!(rownames(tstat_150and500) %in% proteins_150only), ]

        ## Cluster proteins based on the correlation between their t-stat vectors
        dst <- sqrt(2 * (1 - cor(t(tstat_150and500), use = "pairwise.complete")))
        dst[is.na(dst)] <- sqrt(2 * (1 - (-1)))
        clst <- hclust(as.dist(dst))

        ## Reorder proteins by the clustering
        protorder <- c(rownames(tstat_150and500)[clst$order], proteins_150only)
        tstat <- tstat[protorder, ]

        ## Reorder experiments to keep a "diagonal" in the heatmap
        tstat <- reorderColsByRows(tstat)

        tmpbaits <- .getOrigBaitNameFromSimplifiedComparison(colnames(tstat), 
                                                             idmap = idmap)
        colsplit <- ifelse(
            tmpbaits %in% baitclass$Gene_name[baitclass$class == "nobait"], 
            "No\nbait", 
            ifelse(
                tmpbaits %in% baitclass$Gene_name[baitclass$class == "150only"],
                "150 mM NaCl",
                ifelse(
                    tmpbaits %in% baitclass$Gene_name[baitclass$class == "150and500"],
                    "150 and\n500 mM NaCl", "NA"
                )
            )
        )
    } else if (conc == 500) {
        tstat_onlybait <- 
            tstat[, colSums(tstats$tstatsfilt > 0) == 1 |
                      (.getOrigBaitNameFromComparison(colnames(tstats$tstatsfilt), 
                                                      idmap = idmap) %in% 
                      baitclass$Gene_name[baitclass$class500 == "Lost in 500 mM"])]
        tstat_several <- 
            tstat[, colSums(tstats$tstatsfilt > 0) > 1 & 
                      (.getOrigBaitNameFromComparison(colnames(tstats$tstatsfilt), 
                                                      idmap = idmap) %in% 
                      baitclass$Gene_name[baitclass$class500 != "Lost in 500 mM"])]
        tstat_onlybait <- tstat_onlybait[, colnames(tstat_onlybait) != "Untagged_plate"]
        tstat_several <- tstat_several[, colnames(tstat_several) != "Untagged_plate"]

        proteins_onlybait <-
            .getProteinNameFromSimplifiedComparison(colnames(tstat_onlybait))
        stopifnot(all(proteins_onlybait %in% rownames(tstat)))
        tstat_several <- tstat_several[!(rownames(tstat_several) %in% proteins_onlybait), ]

        ## Cluster proteins based on the correlation between their t-stat vectors
        dst <- sqrt(2 * (1 - cor(t(tstat_several), use = "pairwise.complete")))
        dst[is.na(dst)] <- sqrt(2 * (1 - (-1)))
        clst <- hclust(as.dist(dst))

        ## Reorder proteins by the clustering
        protorder <- c(rownames(tstat_several)[clst$order], proteins_onlybait)
        tstat <- tstat[protorder, ]

        ## Reorder conditions to keep a "diagonal" in the heatmap
        tstat <- reorderColsByRows(tstat)

        colsplit <- factor(
            structure(baitclass$class500[
                match(.getOrigBaitNameFromSimplifiedComparison(colnames(tstat), 
                                                               idmap = idmap), 
                      baitclass$Gene_name)],
                names = colnames(tstat)),
            levels = c("Retained in 500 mM", "Lost in 500 mM", ""))
        levels(colsplit) <- c("Interactions\nin 500 mM NaCl",
                              "No interactions\nin 500 mM NaCl", "")
    } else {
        stop("Unknown concentration")
    }
    return(list(tstat = tstat, colsplit = colsplit, nbrIntMat = nbrIntMat))
}

# for each region in `from`, calculate the distance to its nearest element in `to`
# if there are no element in `to` on the sequence of `from`, set distance to NA
calcDistanceToNearest <- function(from, to) {
    stopifnot(exprs = {
        is(from, "GRanges")
        is(to, "GRanges")
    })
    tmp <- distanceToNearest(x = from, subject = to)
    d <- rep(NA, length(from))
    d[queryHits(tmp)] <- mcols(tmp)$distance
    return(d)
}
```

@@

@@

# create main heatmap with annotations
hm3 <- Heatmap(m[sel_peaksBin, ],
               col = circlize::colorRamp2(
                   breaks = seq(0, max(m), length.out = 64),
                   colors = colorRampPalette(binary_heatmap_colors[c("FALSE", "TRUE")])(64)),
               border = TRUE, border_gp = gpar(lwd = 0.5),
               cluster_rows = TRUE, row_dend_width = unit(25, "mm"), show_row_dend = FALSE,
               cluster_columns = TRUE, column_dend_height = unit(10, "mm"),
               row_title_gp = gpar(fontsize = fl),
               row_split = peaktype[sel_peaksBin], cluster_row_slices = FALSE,
               left_annotation = peakAnnot,
               right_annotation = fracTFsAnnot,
               top_annotation = sampleAnnot,
               bottom_annotation = sampleLabels,
               show_row_names = FALSE, show_column_names = FALSE,
               use_raster = TRUE, show_heatmap_legend = FALSE,
               width = unit(w_peaksHm, "mm"), # heatmap body width
               heatmap_height = unit(h_peaksHm, "mm")) # whole heatmap height

@@

# create main heatmap with annotations
hm4 <- Heatmap(m[sel_peaksEnr, ], name = "log2 IP/input",
               col = circlize::colorRamp2(
                   breaks = c(-mx, seq(-qs, qs, length.out = 62), mx),
                   colors = colorRampPalette(enrichment_heatmap_colors)(64)),
               cluster_rows = FALSE,
               cluster_columns = column_dend(hm3), column_dend_height = unit(10, "mm"),
               row_title_gp = gpar(fontsize = fl),
               row_split = peaktype[sel_peaksEnr], cluster_row_slices = FALSE,
               border = TRUE, border_gp = gpar(lwd = 0.5),
               left_annotation = peakAnnot,
               right_annotation = peakAnnot2,
               top_annotation = sampleAnnot,
               bottom_annotation = sampleLabels,
               show_row_names = FALSE, show_column_names = FALSE,
               use_raster = TRUE, show_heatmap_legend = TRUE,
               heatmap_legend_param = list(at = round(c(-mx, 0, mx), 1),
                                           labels_gp = gpar(fontsize = fs),
                                           border = "gray10",
                                           legend_direction = "horizontal",
                                           legend_width = unit(30, "mm"),
                                           title = "TF log2 IP/input",
                                           title_position = "topcenter",
                                           title_gp = gpar(fontsize = fl)),
               width = unit(w_peaksHm, "mm"), # heatmap body width
               heatmap_height = unit(h_peaksHm, "mm"))

@@

```{r plot-chipReplPairs}
#| fig.width: 9
#| fig.height: 9

@@

(gg_enrpairs <- plot_grid(plotlist = pL, nrow = 9, ncol = 9))
```

@@

cowplot::plot_grid(
    hm_peaksEnr,
    hm_peaksBin,
    hm_150,
    hm_500,
    nrow = 2,
    labels = c("A  ChIP", "B  ChIP", "C  IP-MS", "D  IP-MS"),
    align = "v",
    axis = "b"
)
```

@@

```{r suppFig2}
#| fig.width: 11
#| fig.height: 11
#| warning: false

cowplot::plot_grid(
    gg_enrpairs,
    scale = 0.9,
    labels = "A"
)
```

>>>>

---
title: "Figure 2 - Overview heatmaps for IP-MS and ChIP"
author: "Charlotte Soneson, Michael Stadler, Merle Skribbe"
date: "`r Sys.Date()`"
output: 
    html_document:
        theme: united
        toc: true
        toc_float: true
        code_folding: hide
params:
    rds150: "data/ipms_150_sce.rds"
    rds500: "data/ipms_500_sce.rds"
    complexes: "data/complexes.json"
    idmap: "data/id_mapping_table.txt"
    baitclass: "data/ipms_bait_class.txt"
    dbdtxt: "data/TF_DBD_annotation.txt"
    peakcsv: "data/fused_peaks_filtered.csv.gz"
    peakenr: "data/fused_peaks_filtered_enrichments.csv.gz"
    peakatacfile: "data/fused_peaks_filtered_counts-external_atac.csv.gz"
    peakchipfile: "data/fused_peaks_filtered_counts-external_chip.csv.gz"
    gtf: "reference/Schizosaccharomyces_pombe.ASM294v2.55.gtf"
    ipms150untag: "data/ipms_150_untag_sce.rds"
---

@@

## Source scripts with required helper functions and settings
source("params/plot_settings.R")
source("params/get_testres_function.R")
source("params/ipms_params.R")
source("params/mapping_functions.R")
source("params/ipms_heatmap_functions.R")

@@

```{r heatmap-helper}
# for each region in `from`, calculate the distance to its nearest element in `to`
# if there are no element in `to` on the sequence of `from`, set distance to NA
calcDistanceToNearest <- function(from, to) {
    stopifnot(exprs = {
        is(from, "GRanges")
        is(to, "GRanges")
    })
    tmp <- distanceToNearest(x = from, subject = to)
    d <- rep(NA, length(from))
    d[queryHits(tmp)] <- mcols(tmp)$distance
    return(d)
}
```

@@

## IP-MS data for HOT TFs

```{r ipms-hot}
#| fig.width: 9
#| fig.height: 9

sce150untag <- readRDS(params$ipms150untag)

res150untag <- .getTestCols(sce150untag, adjp_cutoff = 0.01, logfc_cutoff = 1)
tstat <- res150untag$tstat
colnames(tstat) <- .getProteinNameFromComparison(colnames(tstat))
tstat <- tstat[rownames(tstat) %in% colnames(tstat), ]
tstat[is.na(tstat)] <- 0
dim(tstat)

hot_tfs <- c("Php3", "Sak1", "Pcr1", "Prr1", "Atf1", "Rst2", "Adn2", 
             "Adn3", "Hsr1", "Phx1", "Pho7")
stopifnot(all(hot_tfs %in% rownames(tstat)))
stopifnot(all(hot_tfs %in% colnames(tstat)))
stopifnot(all(rownames(tstat) %in% colnames(tstat)))
tstat <- tstat[c(hot_tfs, setdiff(rownames(tstat), hot_tfs)), ]
tstat <- tstat[, c(rownames(tstat), setdiff(colnames(tstat), rownames(tstat)))]

hmipmshot <- Heatmap(
    tstat, 
    col = circlize::colorRamp2(
        breaks = seq(3.0, 13.0, length.out = 64),
        colors = colorRampPalette(binary_heatmap_colors[c("FALSE", "TRUE")])(64)),
    border = TRUE,
    row_split = ifelse(rownames(tstat) %in% hot_tfs, "HOT TFs", "Other TFs"), 
    column_split = ifelse(colnames(tstat) %in% hot_tfs, "HOT TFs", "Other TFs"),
    cluster_rows = FALSE, cluster_columns = FALSE,
    row_title_gp = gpar(fontsize = fl), row_names_gp = gpar(fontsize = fs),
    column_title_gp = gpar(fontsize = fl), show_column_names = FALSE,
    heatmap_legend_param = list(labels_gp = gpar(fontsize = fs),
                                border = "gray10",
                                legend_direction = "horizontal",
                                legend_width = unit(24, "mm"),
                                title_position = "lefttop",
                                title_gp = gpar(fontsize = fl)),
    na_col = bg_color,
    name = "Mod. t-stat.")

hm_ipms_hot <- grid.grabExpr(
    hmipmshot <- draw(hmipmshot, merge_legend = TRUE, 
                      heatmap_legend_side = "bottom")
)
plot_grid(hm_ipms_hot)
```

@@

# create main heatmap with annotations
hm3 <- Heatmap(m[sel_peaksBin, ],
               column_title = "ChIP (binary)",
               column_title_side = "top",
               column_title_gp = gpar(fontsize = ft),
               col = circlize::colorRamp2(
                   breaks = seq(0, max(m), length.out = 64),
                   colors = colorRampPalette(binary_heatmap_colors[c("FALSE", "TRUE")])(64)),
               border = TRUE, border_gp = gpar(lwd = 0.5),
               cluster_rows = TRUE, row_dend_width = unit(25, "mm"), show_row_dend = FALSE,
               cluster_columns = TRUE, column_dend_height = unit(10, "mm"),
               row_title_gp = gpar(fontsize = fl),
               row_split = peaktype[sel_peaksBin], cluster_row_slices = FALSE,
               left_annotation = peakAnnot,
               right_annotation = fracTFsAnnot,
               top_annotation = sampleAnnot,
               bottom_annotation = sampleLabels,
               show_row_names = FALSE, show_column_names = FALSE,
               use_raster = TRUE, show_heatmap_legend = FALSE,
               width = unit(w_peaksHm, "mm"), # heatmap body width
               heatmap_height = unit(h_peaksHm, "mm")) # whole heatmap height

@@

# create main heatmap with annotations
hm4 <- Heatmap(m[sel_peaksEnr, ], name = "log2 IP/input",
               column_title = "ChIP (enrichments)",
               column_title_side = "top",
               column_title_gp = gpar(fontsize = ft),
               col = circlize::colorRamp2(
                   breaks = c(-mx, seq(-qs, qs, length.out = 62), mx),
                   colors = colorRampPalette(enrichment_heatmap_colors)(64)),
               cluster_rows = FALSE,
               cluster_columns = column_dend(hm3), column_dend_height = unit(10, "mm"),
               row_title_gp = gpar(fontsize = fl),
               row_split = peaktype[sel_peaksEnr], cluster_row_slices = FALSE,
               border = TRUE, border_gp = gpar(lwd = 0.5),
               left_annotation = peakAnnot,
               right_annotation = peakAnnot2,
               top_annotation = sampleAnnot,
               bottom_annotation = sampleLabels,
               show_row_names = FALSE, show_column_names = FALSE,
               use_raster = TRUE, show_heatmap_legend = TRUE,
               heatmap_legend_param = list(at = round(c(-mx, 0, mx), 1),
                                           labels_gp = gpar(fontsize = fs),
                                           border = "gray10",
                                           legend_direction = "horizontal",
                                           legend_width = unit(30, "mm"),
                                           title = "TF log2 IP/input",
                                           title_position = "topcenter",
                                           title_gp = gpar(fontsize = fl)),
               width = unit(w_peaksHm, "mm"), # heatmap body width
               heatmap_height = unit(h_peaksHm, "mm"))

@@

```{r plot-chipReplPairs}
#| fig.width: 9
#| fig.height: 18

@@

(gg_enrpairs <- plot_grid(plotlist = pL, ncol = 6))
```

@@

cowplot::plot_grid(
    hm_peaksEnr,
    hm_peaksBin,
    hm_150,
    hm_500,
    nrow = 2,
    labels = c("A", "B", "C", "D"),
    align = "v",
    axis = "b"
)
```

@@

```{r suppFig2}
#| fig.width: 11
#| fig.height: 18
#| warning: false

cowplot::plot_grid(
    gg_enrpairs,
    hm_ipms_hot,
    rel_widths = c(1, 0.65), 
    scale = 0.9, nrow = 1,
    labels = c("A", "B")
)
```
